\scshape Evolutionary dynamics, equilibrium selection, and what population and quantitative genetics have taught evolutionary game theory

### 2022-10-10 manuscript.tex
moar writing

%%% new commands and macros
\newcommand{\der}{\mathop{}\!\mathrm{d}}
\newcommand{\Oh}[1]{O\left(#1\right)}
\newcommand{\ExpectationOperator}{\mathrm{E}}
\NewDocumentCommand \E {d[] d()} {%
  \IfNoValueTF {#1} {%
    \IfNoValueTF {#2} {%
      \mathop{\kern0pt\ExpectationOperator}\nolimits%
    }{%
      \mathop{\kern0pt\ExpectationOperator}\nolimits\left(#2\right)%
    }}{%
    \IfNoValueTF {#2} {%
      \mathop{\kern0pt\ExpectationOperator#1}\nolimits%
    }{%
      \mathop{\kern0pt\ExpectationOperator#1}\nolimits\left(#2\right)%
    }%
  }
}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}
\newcommand{\Fst}{F_{\mathrm{ST}}}
\newcommand{\mean}[1]{\overline{#1}}

@@

The role of evolutionary game theory and the ESS method specifically in evolutionary biology has been at the crux of some of the most important conceptual debates in the field including the role of natural selection and adaptation vis-a-vis other forces \cite[e.g.,][]{MaynardSmith:1978,Gould:Lewontin:1979,Lewontin:1979,Orzack:Sober:1994,Gardner:2017,Kern:Hahn:2018,Jensen:Payseur:2019} and the importance of kin and group selection relative to individual selection
\cite[e.g.,][]{Maynard-Smith:1964,Hamilton:1963,Price:1972:cov,Wilson:Wilson:2007,Leigh:2010,Akcay:Van-Cleve:2012,West:Griffin:2007,Gardner:Grafen:2009,Nowak:Tarnita:2010,Abbot:Abe:2011,Allen:Nowak:2013,Birch:2014,Birch:2017,Nowak:McAvoy:2017}. These debates might be captured in part by the following two questions: \begin{ilnum} \item \label{q:I} How do ESS models that focus on individual fitness capture the effects of kin selection or group selection? \item \label{q:II} How does an ESS account for mutation, genetic drift, and other evolutionary forces? \end{ilnum} In this work, I will describe how conceptual advances in evolutionary theory since Maynard Smith and Price \cite{Maynard-Smith:Price:1973} have shed light on both of these questions and have led to a more integrative framework for understanding how multiple evolutionary forces in complex populations can lead to diverse phenotypes.

@@

@@

Following the development of the ESS approach by Maynard Smith and Price \cite{Maynard-Smith:Price:1973}, mathematical biologists evaluated the approach in light of the older and established theory population genetics. Population genetic models can explicitly address how both natural selection and genetic mechanisms including mutation, segregation, and recombination affect allele and genotype frequency dynamics over time. These models typically assume that evolution changes the frequencies of a fixed set of genotypes and the emergence of new genotypes through the large-effect mutations (i.e., the creation of new alleles or loci) occurs on a longer timescale than that of the model. For this reason, these population genetic models are often said to represent a ``short-term'' evolutionary process. Population genetic models showed that \cite{Cavalli-Sforza:Feldman:1978,Eshel:1982}

>>>>

%%% new commands and macros
\newcommand{\der}{\mathop{}\!\mathrm{d}}
\newcommand{\Oh}[1]{O\left(#1\right)}
\newcommand{\ExpectationOperator}{\mathrm{E}}
\NewDocumentCommand \E {d[] d()} {%
  \IfNoValueTF {#1} {%
    \IfNoValueTF {#2} {%
      \mathop{\kern0pt\ExpectationOperator}\nolimits%
    }{%
      \mathop{\kern0pt\ExpectationOperator}\nolimits\left(#2\right)%
    }}{%
    \IfNoValueTF {#2} {%
      \mathop{\kern0pt\ExpectationOperator#1}\nolimits%
    }{%
      \mathop{\kern0pt\ExpectationOperator#1}\nolimits\left(#2\right)%
    }%
  }
}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}
\newcommand{\Fst}{F_{\mathrm{ST}}}
\newcommand{\mean}[1]{\overline{#1}}
\newcommand{\w}{w}
\newcommand{\ess}[1]{#1^*}

@@

The role of evolutionary game theory and the ESS method specifically in evolutionary biology has been at the crux of some of the most important conceptual debates in the field including the role of natural selection and adaptation vis-a-vis other forces \cite[e.g.,][]{MaynardSmith:1978,Gould:Lewontin:1979,Lewontin:1979,Orzack:Sober:1994,Gardner:2017,Kern:Hahn:2018,Jensen:Payseur:2019} and the importance of kin and group selection relative to individual selection
\cite[e.g.,][]{Maynard-Smith:1964,Hamilton:1963,Price:1972:cov,Wilson:Wilson:2007,Leigh:2010,Akcay:Van-Cleve:2012,West:Griffin:2007,Gardner:Grafen:2009,Nowak:Tarnita:2010,Abbot:Abe:2011,Allen:Nowak:2013,Birch:2014,Birch:2017,Nowak:McAvoy:2017}. These debates might be captured in part by the following two questions: \begin{ilnum} \item \label{q:I} How do ESS models that focus on individual fitness capture the effects of kin selection or group selection? \item \label{q:II} How does an ESS account for mutation, genetic drift, and other evolutionary forces? \end{ilnum} In this work, I will describe how conceptual advances in evolutionary theory since Maynard Smith and Price \cite{Maynard-Smith:Price:1973} have shed light on both of these questions and have led to a broader understanding of the ESS method and a more integrative framework for understanding how multiple evolutionary forces in complex populations can lead to diverse phenotypes.

@@

%%%
%%% Detailed outline for post intro
%%%

What is ESS? ESS is just phenotypes and fitness and an equilibrium and stability condition. This matches pop gen for simple genetics \cite{Eshel:1982}.

ESS really is a model of a ``long-term evolution'' process of invasion of mutant by resident where fitness determines invasion. This is the ``unbeatable strategy'' of Hamilton.

Long-term evolution depends on a fitness ``measure'' so ``fitness'' isn't just individual fitness. Really, we track the frequency or density of individual alleles (i.e., gene's eye view). This is lineage growth rate.

Lineage growth rate allows us to measure ESS in complex structure populations with kin and group selection. We measure the growth rate of the lineage in all the genetic and demographic environments it finds itself in.

Genetic environments can include other loci linked by recombination. This is the case of the modifier theory for recombination, mutation, migration, etc. Example from Liberman 2011.

Recognizing the importance of genetic parameters in ESS is the opposite of early work to justify ESS which tried to find conditions when ESS predicted same as pop gen but genetic parameters didn't appear.

One way forward to understanding the complex phenotypic evolution including social evolution should involve using ESS theory that incorporates more genetics including multiple loci. This is notably complementary to methods that study trait coevolution, e.g., Mullon and Lehmann.

%%%
%%% Fleshed out outline material
%%%

\section{What is an ESS?}

The definition of an ESS is relatively simple and only involves a measure of fitness, a set of phenotypes, and a stability condition. Let $\w(x,y)$ be the fitness that an individual with phenotype or strategy $x$ obtains when interacting with an individual with phenotype or strategy $y$. The phenotypes can be drawn from a set of discrete or continous set of values $X$ that represent a single trait like body size or propensity to cooperate. Fitness $\w$ for the moment is simply the survival rate (i.e., viability selection) but as we'll see we can define much more general fitness measures. The stability condition \cite{Maynard-Smith:Price:1973,Maynard-Smith:1974} says that a phenotype $\ess{x}$ is an ESS when no alternative strategy $y$ can improve upon the fitness that $\ess{x}$ receives when interacting with itself:
\begin{subequations}
  \label{eq:ess:def}
\begin{equation}
  \label{eq:ess:def:a}
  \w(\ess{x}, \ess{x}) \ge \w(y, \ess{x})
\end{equation}
for all phenotypes $y$. In the case that the strategy $y$ gets the same fitness interacting with $\ess{x}$ as $\ess{x}$ does with itself (equation \eqref{eq:ess:def:a} holds as an equality) then $\ess{x}$ must do better interacting with $y$ than $y$ does with itself:
\begin{equation}
  \label{eq:ess:def:b}
  \w(\ess{x}, y) > \w(y, y)
\end{equation}
\end{subequations}
The two conditions in \eqref{eq:ess:def} are equivalent to requiring that phenotype $\ess{x}$ receives higher average fitness than any alternative phenotype $y$ when $y$ is rare in the population \cite{Maynard-Smith:1974,Bishop:Cannings:1976}.

Notably, the conditions in \eqref{eq:ess:def} do not reference the genetic basis of the trait $X$ nor do they describe how the frequency or mean value of the trait changes over time. Population genetics models, on the other hand, are built to measure the dynamical process of evolution via the combined effects of natural selection and segregation, recombination, mutation, genetic drift, and other evolutionary forces \cite{Crow:Kimura:1970,Ewens:2004}. A natural question then is how the long-run behavior of a population genetic model (e.g., its equilibrium genotypes and induced phenotypes) compares to the equilibrium phenotypes obtained from an analogous ESS model. This question was tackled by mathematical population biologists beginning in the late 1970s and through 1980s and 1990s \cite[e..g,][]{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980,Eshel:1982,Hofbauer:Schuster:1982,Lessard:1984,Cressman:1988,Cressman:Hines:1984,Cressman:Hofbauer:1996,Hammerstein:1996,Weissing:1996,Eshel:1996,Eshel:Feldman:1984}. Early work in the simplest haploid asexual model with discrete strategies showed that the ESSs of obtained from condition \eqref{eq:ess:def} must be stable equilibria of haploid population genetic model \cite{Taylor:Jonker:1978,Hofbauer:Schuster:1979,Zeeman:1980}. This result isn't particularly surprising since natural selection is the only evolutionary force acting in the haploid asexual model. However, this idea was extended further by Ilan Eshel and Sabin Lessard

%%%
%%% old scraps and ramblings
%%%

\section{Stability in dynamical systems and the ESS}

For the behavior or evolutionary ecologist interested in a mathematical model of natural selection, an ESS model offers one main advantage over population genetic models: the ESS model can be formulated entirely in terms of phenotypes and ftiness and can avoid relying explicitly on genetic parameters (e.g., number of loci affecting the trait, recombination rates, etc) because the implicit genetic assumptions (i.e., additive interactions between alleles within a locus and across loci) imply a population genetic model with the same assumptions would generate the same ESSs \cite{Eshel:1982}.

@@

Following the development of the ESS approach by Maynard Smith and Price \cite{Maynard-Smith:Price:1973}, mathematical biologists evaluated the approach in light of the older and established theory population genetics. Population genetic models can explicitly address how both natural selection and genetic mechanisms including mutation, segregation, and recombination affect allele and genotype frequency dynamics over time. These models typically assume that evolution changes the frequencies of a fixed set of genotypes and that the emergence of new genotypes through the large-effect mutations (i.e., the creation of new alleles and loci) occurs on a longer timescale than that of the model. For this reason, these population genetic models are often said to represent a ``short-term'' evolutionary process. Population genetic models showed that \cite{Cavalli-Sforza:Feldman:1978,Eshel:1982}
